Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-26 paper/main.tex
edits to 4.3

Here we show prediction results for $N=6$ held-out cosmologies for the Mira-Titan dataset. 
We compare our method with Cosmic Emu \citep{moran2023mira}, the state-of-the-art emulator 
constructed on the same $m=111$ Mira-Titan cosmologies. Cosmic Emu uses a Bayesian approach 
where the ``true'' spectrum for each cosmology is modeled with a process convolution on Brownian motion. 
Nonstationarity is permitted through modeling the bandwidth parameter with a process 
convolution as well; this composition is referred to as a deep process convolution 
\citep[DPC,][]{moran2023mira}. Cosmic Emu also uses a PC approach to predict power matter 
spectra for held-out cosmologies as a function of $\psi$ \citep{gattiker2020sepia}.

Using our proposed DGP.FCO approach and Cosmic Emu, we obtain two sets of predictions for $\hat{S}_t(\psi_t)$ 
with $t\in\{1,\ldots,6\}$ indexing the six held-out cosmologies.  In this real-world example, we have
no ``truth" for the Mira-Titan hold-outs against which to benchmark. In an effort to assess the 
accuracy of these predictions, we also separately fit the DGP.FCO model on each of these hold outs
and use the corresponding posterior means as a reference. The individual predictions for each method and 
cosmology are shown in Figure \ref{fig:plot_pred_1to6}, where the posterior mean is subtracted so that
predictions lines closer to zero represent a better fit.

@@

From these predictions, we can calculate MSE when treating the trained DGP.FCO predictions as a benchmark. 
Our DGP.FCO prediction achieves the greatest reduction in MSE in the region where perturbation theory
and the low-resolution runs are deemed unbiased (see Figure \ref{fig:mse_by_k}). For the region where only 
the high-resolution run is unbiased, Cosmic Emu generally has a slightly lower average RMSE. Our average MSE 
is lower at 55\% of $k$ values considered. These results indicate that DGP.FCO competes favorably with the 
state-of-the-art Cosmic Emu model.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mse_by_k.jpeg}
    \caption{Average MSE across each of the 6 hold out cosmologies, for each $k$ value. Cosmic Emu 
    predictions are shown in dotted blue, and DGP.FCO predictions are shown in dashed orange.}
    \label{fig:mse_by_k}
\end{figure}

>>>>

Here we show prediction results for $N=6$ held-out cosmologies for the Mira-Titan dataset. 
We compare our method with Cosmic Emu \citep{moran2023mira}, the state-of-the-art emulator 
constructed on the same $m=111$ Mira-Titan cosmologies. Cosmic Emu uses a Bayesian approach 
where the ``true'' spectrum for each cosmology is modeled with a process convolution on Brownian motion. 
Nonstationarity is permitted through modeling the bandwidth parameter with a process 
convolution as well; this composition is referred to as a deep process convolution 
\citep[DPC;][]{moran2023mira}. Cosmic Emu also uses a principal components model to predict power matter 
spectra for held-out cosmologies as a function of $\psi$ \citep{gattiker2020sepia}.

Using our proposed DGP.FCO approach and Cosmic Emu, we obtain two sets of predictions for $\hat{S}_t(\psi_t)$ 
with $t\in\{1,\ldots,6\}$ indexing the six held-out cosmologies.  In this real-world example, we have
no ``truth" for the Mira-Titan hold-outs against which to benchmark. In an effort to assess the 
accuracy of these predictions, we also separately fit our DGP.FCO model on each of these testing cosmologies
(leveraging the perturbation theory, low resolution runs, and high resolution run to predict the underlying
power matter spectrum).  Although these predicted ``in-sample'' spectra are not formal ``truths,'' we find them to be
useful benchmarks for our ``out-of-sample'' predictions.  The individual predictions for each method and 
held-out cosmology are shown in Figure \ref{fig:plot_pred_1to6}, centered by the in-sample DGP.FCO
posterior means.  Predictions closer to zero represent a better fit.

@@

From these predictions, we can calculate MSE against the in-sample DGP.FCO predictions across the 6
test cosmologies. 
Our out-of-sample DGP.FCO + PC prediction achieves the greatest reduction in MSE in the region 
where perturbation theory and the low-resolution runs are deemed unbiased (see Figure \ref{fig:mse_by_k}). 
For the region where only the high-resolution run is unbiased, Cosmic Emu has a marginally lower
MSE. Our MSE is lower at 55\% of $k$ values considered. These results indicate that DGP.FCO 
competes favorably with the state-of-the-art Cosmic Emu model.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mse_by_k.jpeg}
    \caption{MSE across each of the 6 hold out cosmologies for each $k$ value. Cosmic Emu 
    performance is shown in dotted blue, and DGP.FCO performance is shown in dashed orange.}
    \label{fig:mse_by_k}
\end{figure}
